---
title: "RNA-seq of WT vs helD knock-out in Mycobacterium smegmatis"
author: "Martin Převorovský, prevorov@natur.cuni.cz, http://www.GenoMik.cz"
output: html_document
---

The main aims of this experiment are to:

* identify genes whose expression levels change in response to presence/absence of HelD in both exponential and stationary phase of growth

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rtracklayer)
library(GenomicAlignments)
library(GenomicFeatures)
library(DESeq2)
library(RColorBrewer)
library(pheatmap)

genome_dir <- './genome_Msmegmatis/'
genome_annot <- paste0(genome_dir, 'GCF_000015005.1_ASM1500v1_genomic.customized.gff')
bam_dir <- './BAM_RNA-seq_Msmegmatis/'
coverage_dir <- './coverage_RNA-seq_Msmegmatis/'
deg_dir <- './DEG_RNA-seq_Msmegmatis/'
image_dir <- './images_Msmegmatis/'
```

Variables for shell scripts are stored in the file "shell_variables_RNA-seq_Msmegmatis.sh" and can be modified as needed:

```{bash list_variables}
cat shell_variables_RNA-seq_Msmegmatis.sh
```


# *Mycobacterium smegmatis* MC2 155 genome sequence and annotation

* genome sequence used: gi|118467340|ref|NC_008596.1|

* genome annotation used: genome-build-accession NCBI_Assembly:GCF_000015005.1

* GFF annotation was edited to include the Ms1 ncRNA and the rnpB genes

* spike sequences added to the genome (the 4-spike mix was used)

```{bash, message=FALSE, eval=TRUE}
source shell_variables_RNA-seq_Msmegmatis.sh

zcat "${genome_dir}GCF_000015005.1_ASM1500v1_genomic.gff.gz" > "${genome_annot}"
sed -i 's/NC_008596.1/gi|118467340|ref|NC_008596.1|/g' "${genome_annot}"
echo -e "gi|118467340|ref|NC_008596.1|\tcustom\tgene\t4392642\t4392993\t.\t-\t.\tID=gene99998;locus_tag=rnpB;gene=rnpB;Name=MSMEG_rnpB" >> "${genome_annot}"
echo -e "gi|118467340|ref|NC_008596.1|\tcustom\tCDS\t4392642\t4392993\t.\t-\t0\tID=cds99998;Parent=gene99998;locus_tag=rnpB" >> "${genome_annot}"
echo -e "gi|118467340|ref|NC_008596.1|\tcustom\tgene\t6242368\t6242672\t.\t+\t.\tID=gene99999;locus_tag=Ms1;gene=Ms1;Name=MSMEG_Ms1" >> "${genome_annot}"
echo -e "gi|118467340|ref|NC_008596.1|\tcustom\tCDS\t6242368\t6242672\t.\t+\t0\tID=cds99999;Parent=gene99999;locus_tag=Ms1" >> "${genome_annot}"

cat "${genome/\.fasta/_no_spikes\.fasta}" ./spikes_4.fasta > "${genome}"

samtools faidx "${genome}"
cut -f1,2 "${genome}.fai" > "${genome_dir}chrom.sizes"
```


# QC of raw sequence reads

* Libraries were sequenced at the Institute of Molecular Genetics, Czech Academy of Sciences, Prague (Illumina NextSeq 550).

* Technical quality of library construction and sequencing run was tested using `fastQC`.

```{bash}
fastqc --version
multiqc --version
```

```{bash, eval=TRUE}
source shell_variables_RNA-seq_Msmegmatis.sh

mkdir -p "${QC_dir}"
mapfile -t fastq_files < <(ls -1 "${fastq_dir}" | grep "${fastq_file_ext}")
for i in ${fastq_files[@]};
do
  fastqc --quiet --outdir "${QC_dir}" --threads "${CPU}" "${fastq_dir}${i}"
done

multiqc -f -m fastqc -i RNA-seq_Msmegmatis -o "${QC_dir}" "${QC_dir}"
```

Results of QC:

* 86 nt SE reads

* 5.6-11.3 M reads per sample

* no adapter contamination

* all samples give error on per base sequence content (due to random priming)

* all samples give error/warning on per sequence GC content (could be due to extremely high expression of some RNAs, such as Ms1 in stationary phase). 

!!! Sample 6 (helD_KO EX2) should be from exponential phase, but the GC profile resembles rather stationary-phase samples

* all samples give error on high read duplication levels (normal and OK for transcriptome analysis)

* all samples give error/warning on overrepresented sequences

* read quality scores somewhat deteriorate at the end → 3' end trimming of reads is needed


# Read trimming

* trim parts of the reads where average base quality in a 4 nt window drops below 20

* only keep trimmed reads that are at least 20 nt long

* adapter trimming not needed

```{bash}
java -jar /opt/Trimmomatic-0.39/trimmomatic-0.39.jar -version
```

```{bash, eval=TRUE}
source shell_variables_RNA-seq_Msmegmatis.sh

mkdir -p "${fastq_trim_dir}"
mapfile -t fastq_files < <(ls -1 "${fastq_dir}" | grep "${fastq_file_ext}")
for i in ${fastq_files[@]};
do
  java -jar /opt/Trimmomatic-0.39/trimmomatic-0.39.jar SE -phred33 -threads "${CPU}" "${fastq_dir}${i}" "${fastq_trim_dir}${i}.trim.fastq.gz" SLIDINGWINDOW:4:20 MINLEN:20
done
```


# QC of quality-trimmed sequence reads

Effects of trimming were determined by `fastQC`.

```{bash, eval=TRUE}
source shell_variables_RNA-seq_Msmegmatis.sh

mkdir -p "${QC_trim_dir}"
mapfile -t fastq_files < <(ls -1 "${fastq_trim_dir}" | grep "${fastq_file_ext}")
for i in ${fastq_files[@]};
do
  fastqc --quiet --outdir "${QC_trim_dir}" --threads "${CPU}" "${fastq_trim_dir}${i}"
done

multiqc -f -m fastqc -i RNA-seq_Msmegmatis_trimmed -o "${QC_trim_dir}" "${QC_trim_dir}"
```

Results of QC:

* dropped ~2-3% of reads

* 20-86 nt reads (vast majority of reads are >82 nt)

* read quality scores now OK


# Read mapping into *M. smegmatis* genome

```{bash}
hisat2 --version
samtools --version
```

```{bash, eval=TRUE}
source shell_variables_RNA-seq_Msmegmatis.sh

# build HISAT2 index
hisat2-build --quiet -p "${CPU}" "${genome}" "${genome}"

# map
mkdir -p "${bam_dir}"
mapfile -t fastq_files < <(ls -1 "${fastq_trim_dir}" | grep "${fastq_file_ext}")
for i in ${fastq_files[@]};
do
	infile="${fastq_trim_dir}${i}"
	outfile="${bam_dir}${i}.bam"
  echo "${outfile}"
	hisat2 -x "${genome}" -U "${infile}" --threads "${CPU}" --no-spliced-alignment -k 10 --summary-file "${outfile}.log" | samtools view -b -F 256 --threads "${CPU}" - | samtools sort --threads "${CPU}" -o "${outfile}" - 
	samtools index "${outfile}"
	samtools view -c --threads "${CPU}" "${outfile}"
done
```

Notes

* To properly map reads coming from repeat regions (e.g., rRNA), HISAT was set to search for up to 10 matching positions, then only the best (marked as primary by HISAT) alignment was kept.

	* **The kit used for sequencing library preparation produces reads that are reverse-complement to the original RNA sequence. Therefore, reads map to the opposite strand compared to the corresponding gene's orientation. This needs to be accounted for during count table generation (reads are reversed) and generation of strand-specific BAM and BigWig files (plus-strand reads are used for computing real minus-strand coverage, and vice versa).**
	
* Mapping rate >92%.


# Create genome coverage files for each sample 

* Note that strand polarity is swapped during the calculation (sequencing libraries are reverse-complement of the original RNA).

```{bash}
deeptools --version
```

```{bash, eval=TRUE}
source shell_variables_RNA-seq_Msmegmatis.sh

mkdir -p "${coverage_dir}"
mapfile -t bam_files < <(ls -1 "${bam_dir}" | grep "\.bam$")
for i in ${!bam_files[@]};
do
  scaling_factor=`samtools idxstats "${bam_dir}${bam_files[${i}]}" | head -n 5 | cut -f3 | paste -s -d+ | bc`
  scaling_factor=`echo "scale=6; 1/(${scaling_factor}/1000000)" | bc`
	input_file="${bam_dir}${bam_files[${i}]}"
	bamCoverage --binSize "${bin_size}" -p "${CPU}" --bam "${input_file}" -o "${coverage_dir}${bam_files[${i}]}.bw" --scaleFactor "${scaling_factor}"
	bamCoverage --binSize "${bin_size}" -p "${CPU}" --bam "${input_file}" -o "${coverage_dir}${bam_files[${i}]}.plus.bw" --scaleFactor "${scaling_factor}" --samFlagInclude 16
	scaling_factor=`echo "${scaling_factor}*-1" | bc`
	bamCoverage --binSize "${bin_size}" -p "${CPU}" --bam "${input_file}" -o "${coverage_dir}${bam_files[${i}]}.minus.bw" --scaleFactor "${scaling_factor}" --samFlagExclude 16
done
```

Notes

* Checked all samples visually in IGV. Sample 6 (helD_KO EX2) seems to actually be WT from STAT phase → sample will be excluded!

* Most reads come from the MSMEG_2093 (rna10) locus. Many reads from Ms1 and rnpB loci.

* rRNA reads clearly present, but quantity looks reasonable.


# Determine rRNA content

```{bash, eval=TRUE}
source shell_variables_RNA-seq_Msmegmatis.sh

rRNA_1='gi|118467340|ref|NC_008596.1|:3819833-3825494'
rRNA_2='gi|118467340|ref|NC_008596.1|:5024129-5031105'

bam_files=( $(ls -1 -d "${bam_dir}"* | grep "\.bam$") )

echo -e "file\tall\tmapped\tunmapped\trRNA\tother_mapped" > rRNA_read_counts_Msmegmatis
for i in ${bam_files[@]};
do
  all_reads=`samtools view -c "${i}"`
  mapped_reads=`samtools view -c -F 4 "${i}"`
  unmapped_reads=`samtools view -c -f 4 "${i}"`
  rRNA_reads=`samtools view -c "${i}" "${rRNA_1}" "${rRNA_2}"`
  other_mapped_reads=`echo "${mapped_reads}-${rRNA_reads}" | bc`
  echo -e "${i/$bam_dir/}\t${all_reads}\t${mapped_reads}\t${unmapped_reads}\t${rRNA_reads}\t${other_mapped_reads}" >> rRNA_read_counts_Msmegmatis
done
```


# Determine spike-in content

```{bash, eval=TRUE}
source shell_variables_RNA-seq_Msmegmatis.sh

spikes=("Plat" "Mos" "Elav2" "Nluc")

bam_files=( $(ls -1 -d "${bam_dir}"* | grep "\.bam$") )

echo -e "file\tall\tmapped\tunmapped\tspikes\tother_mapped" > spike-in_read_counts_Msmegmatis
for i in ${bam_files[@]};
do
  all_reads=`samtools view -c "${i}"`
  mapped_reads=`samtools view -c -F 4 "${i}"`
  unmapped_reads=`samtools view -c -f 4 "${i}"`
  spike_reads=`samtools view -c "${i}" "${spikes[@]}"`
  other_mapped_reads=`echo "${mapped_reads}-${spike_reads}" | bc`
  echo -e "${i/$bam_dir/}\t${all_reads}\t${mapped_reads}\t${unmapped_reads}\t${spike_reads}\t${other_mapped_reads}" >> spike-in_read_counts_Msmegmatis
done
```


# Create count tables of all genes

Due to uncertainty about sample identity, the whole replicate 2 was removed from further analyses.

```{r, eval=TRUE}
gff <- import.gff3(genome_annot)
gff.genes <- gff[gff$type == 'gene', ]

sample_info <- read.delim('sampleInfo', stringsAsFactors = FALSE, header = TRUE, sep = '\t')
sample_info[, 'cultivation_run'] <- factor(sample_info[, 'cultivation_run'])
sample_info[, 'genotype'] <- factor(sample_info[, 'genotype'])
sample_info[, 'phase'] <- factor(sample_info[, 'phase'])
sample_info <- sample_info[sample_info$cultivation_run != 2, ] # remove replicate 2

bam <- BamFileList(list.files(bam_dir, 
                              pattern = '\\.bam$', 
                              full.names = TRUE),
                   asMates = FALSE,
                   yieldSize = 1000000)
bam <- bam[grep('_2', names(bam), invert = TRUE)] # remove replicate 2
names(bam) <- sample_info$sample_ID

se <- summarizeOverlaps(features = gff.genes,
                        reads = bam,
                        mode = 'Union',
                        singleEnd = TRUE,
                        ignore.strand = FALSE,
                        preprocess.reads = invertStrand)
rownames(se) <- gff.genes$locus_tag
colData(se) <- cbind(colData(se), sample_info)
save(file = 'se_RNA-seq_Msmegmatis.rda', se)
#load('se_RNA-seq_Msmegmatis.rda')

# normalize to get TPM values (transcripts per million)
counts <- assay(se)
gene_widths <- width(gff.genes)
counts.norm <- counts
for (i in 1:nrow(counts.norm)){
  counts.norm[i, ] <- counts.norm[i, ] / gene_widths[i]
}
count_sums <- colSums(counts.norm) / 1000000
for (i in 1:length(count_sums)){
  counts.norm[, i] <- counts.norm[, i] / count_sums[i]
}

counts <- cbind(gff.genes$Name, counts)
counts.norm <- cbind(gff.genes$Name, counts.norm)
colnames(counts)[1] <- 'gene_name'
colnames(counts.norm)[1] <- 'gene_name'
write.csv(counts, file ='counts_RNA-seq_Msmegmatis.csv', quote = FALSE)
write.csv(counts.norm, file ='counts_RNA-seq.TPM-normalized_Msmegmatis.csv', quote = FALSE)
rm(gene_widths, count_sums)

# diagnostic plots
pdf(paste0(image_dir, 'RNA-seq_diagnostic_plots.pdf'))
plot(counts.norm[, 'WT_EXP_1'], counts.norm[, 'helD_KO_EXP_1'], log = 'xy', pch = 20, col = rgb(0, 0, 0, 0.2))
points(counts.norm['MSMEG_1367', 'WT_EXP_1'], counts.norm['MSMEG_1367', 'helD_KO_EXP_1'], pch = 20, col = 'red')
points(counts.norm['MSMEG_2758', 'WT_EXP_1'], counts.norm['MSMEG_2758', 'helD_KO_EXP_1'], pch = 20, col = 'green')
points(counts.norm['Ms1', 'WT_EXP_1'], counts.norm['Ms1', 'helD_KO_EXP_1'], pch = 20, col = 'blue')
legend('topleft', c('rpoB', 'sigA', 'Ms1'), col = c('red', 'green', 'blue'), pch = 20)

plot(counts.norm[, 'WT_EXP_3'], counts.norm[, 'helD_KO_EXP_3'], log = 'xy', pch = 20, col = rgb(0, 0, 0, 0.2))
points(counts.norm['MSMEG_1367', 'WT_EXP_3'], counts.norm['MSMEG_1367', 'helD_KO_EXP_3'], pch = 20, col = 'red')
points(counts.norm['MSMEG_2758', 'WT_EXP_3'], counts.norm['MSMEG_2758', 'helD_KO_EXP_3'], pch = 20, col = 'green')
points(counts.norm['Ms1', 'WT_EXP_3'], counts.norm['Ms1', 'helD_KO_EXP_3'], pch = 20, col = 'blue')
legend('topleft', c('rpoB', 'sigA', 'Ms1'), col = c('red', 'green', 'blue'), pch = 20)

plot(counts.norm[, 'WT_STAT_1'], counts.norm[, 'helD_KO_STAT_1'], log = 'xy', pch = 20, col = rgb(0, 0, 0, 0.2))
points(counts.norm['MSMEG_1367', 'WT_STAT_1'], counts.norm['MSMEG_1367', 'helD_KO_STAT_1'], pch = 20, col = 'red')
points(counts.norm['MSMEG_2758', 'WT_STAT_1'], counts.norm['MSMEG_2758', 'helD_KO_STAT_1'], pch = 20, col = 'green')
points(counts.norm['Ms1', 'WT_STAT_1'], counts.norm['Ms1', 'helD_KO_STAT_1'], pch = 20, col = 'blue')
legend('topleft', c('rpoB', 'sigA', 'Ms1'), col = c('red', 'green', 'blue'), pch = 20)

plot(counts.norm[, 'WT_STAT_3'], counts.norm[, 'helD_KO_STAT_3'], log = 'xy', pch = 20, col = rgb(0, 0, 0, 0.2))
points(counts.norm['MSMEG_1367', 'WT_STAT_3'], counts.norm['MSMEG_1367', 'helD_KO_STAT_3'], pch = 20, col = 'red')
points(counts.norm['MSMEG_2758', 'WT_STAT_3'], counts.norm['MSMEG_2758', 'helD_KO_STAT_3'], pch = 20, col = 'green')
points(counts.norm['Ms1', 'WT_STAT_3'], counts.norm['Ms1', 'helD_KO_STAT_3'], pch = 20, col = 'blue')
legend('topleft', c('rpoB', 'sigA', 'Ms1'), col = c('red', 'green', 'blue'), pch = 20)
dev.off()
```


## QC of count table data

```{r, eval=TRUE}
dds <- DESeqDataSet(se, 
                    design = ~ cultivation_run + 
                      genotype +
                      phase)
dds <- DESeq(dds)
rld <- rlog(dds, blind = FALSE)

sampleDists <- dist(t(assay(rld)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- rld$sample_ID
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
pdf(paste0(image_dir, 'heatmap_RNA-seq_sampleDist_rlog_Msmegmatis.pdf'))
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
dev.off()

pdf(paste0(image_dir, 'PCA_RNA-seq_rlog_Msmegmatis.pdf'))
plotPCA(rld, intgroup = c('genotype', 'phase', 'cultivation_run'))
dev.off()
```

* Heatmap results - samples cluster by growth phase, as expected.

* PCA results - STAT samples form one cluster, EXP samples form 2 clusters (batch effects)


# DEG calling

```{r, eval=TRUE}
# Benjamini-Hochberg p value adjustment (FDR)
padj_threshold <- 0.05

dir.create(deg_dir)

# helD-EXP vs WT-EXP
dds <- DESeqDataSet(se, design = ~ cultivation_run + genotype)
dds <- dds[, dds$phase == 'EXP'] # remove STAT samples
dds$phase <- droplevels(dds$phase) # remove factor levels without samples
dds$genotype <- relevel(dds$genotype, ref = 'WT') # change order of factor levels to get helD_KO/WT fold change (= define the reference)
dds <- DESeq(dds)
res <- results(dds, alpha = padj_threshold)
summary(res[res$padj <= padj_threshold & !is.na(res$padj), ])
res <- cbind(counts[, 'gene_name'], res)
colnames(res)[1] <- 'gene_name'
write.csv(res, file = paste0(deg_dir, 'DESeq2results_helD-EXP_vs_WT-EXP.csv'), quote = FALSE)
res.sig <- res[which(res$padj <= padj_threshold), ]
res.sig <- res.sig[order(res.sig$log2FoldChange), ]
write.csv(res.sig, file = paste0(deg_dir, 'DESeq2results_helD-EXP_vs_WT-EXP.SIG.csv'), quote = FALSE)

# helD-STAT vs WT-STAT
dds <- DESeqDataSet(se, design = ~ cultivation_run + genotype)
dds <- dds[, dds$phase == 'STAT'] # remove EXP samples
dds$phase <- droplevels(dds$phase) # remove factor levels without samples
dds$genotype <- relevel(dds$genotype, ref = 'WT') # change order of factor levels to get helD_KO/WT fold change (= define the reference)
dds <- DESeq(dds)
res <- results(dds, alpha = padj_threshold)
summary(res[res$padj <= padj_threshold & !is.na(res$padj), ])
res <- cbind(counts[, 'gene_name'], res)
colnames(res)[1] <- 'gene_name'
write.csv(res, file = paste0(deg_dir, 'DESeq2results_helD-STAT_vs_WT-STAT.csv'), quote = FALSE)
res.sig <- res[which(res$padj <= padj_threshold), ]
res.sig <- res.sig[order(res.sig$log2FoldChange), ]
write.csv(res.sig, file = paste0(deg_dir, 'DESeq2results_helD-STAT_vs_WT-STAT.SIG.csv'), quote = FALSE)
```


***

```{r}
sessionInfo()
```
